### 2023-02-01 tex/semirings.tex
[ writing ] renew generic syntax chapter
[ writing ] more

The rest of this chapter proceeds as follows.
In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
appeared previously in my work with Atkey~\cite{WA21}, and can be seen as a
simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
The idea of \name{} is to augment the simply typed $\lambda$-calculus with
annotations on free variables, and give enough types to manipulate these
annotations.
Given this new calculus $\name$, the first goal is to apply the techniques of
\cref{sec:simple} to it, yielding a simultaneous substitution operation.
To do this, I use \cref{sec:lnd} to introduce notation that allows us to restate
the typing rules of \name{} to not mention contexts explicitly, as was the style
in \cref{sec:simple}.
This new notation --- the \emph{bunched connectives} --- is versatile at
defining simply typed usage-aware syntaxes, and I give further non-$\name$
examples in \cref{sec:rec}.
The application of the techniques of \cref{sec:simple} is left to
\cref{sec:ren-sub-lr}.

\section{A usage-annotated calculus $\name$}\label{sec:lr}
\input{lr.tex}
\section{Bunched connectives}\label{sec:lnd}
\input{lnd.tex}
\section{Additions to and variations of \name{}}\label{sec:variant}
\input{rec.tex}
\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Renaming and substitution of $\name$}\label{sec:ren-sub-lr}

In \cref{sec:semirings}, I defined my calculus of interest $\name$.
In this chapter, I develop its syntactic metatheory up to and including
substitution.
I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
significant changes to the underlying notion of \emph{environment} before doing
so.
I give and informally motivate these changes to environments in
\cref{sec:lrkits}, and prove some properties of the new definition in
\cref{sec:lenv}.
Finally, I apply these new environments to the syntax of $\name$ in
\cref{sec:lrsub} to derive renaming and substitution operators.

\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}

>>>>

The rest of this chapter proceeds as follows.
In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
appeared in my previous work~\citep{WA21}, and can be seen as a
simply typed version of Atkey's dependently typed calculus QTT~\citep{Atkey18}.
The idea of \name{} is to augment the simply typed $\lambda$-calculus with
annotations on free variables, and give enough types to manipulate these
annotations.
Given this new calculus $\name$, the first goal is to apply the techniques of
\cref{sec:simple} to it, yielding a simultaneous substitution operation.
To do this, I use \cref{sec:lnd} to introduce notation that allows us to restate
the typing rules of \name{} to not mention contexts explicitly, as was the style
in \cref{sec:simple}.
This new notation --- the \emph{bunched connectives} --- is versatile at
defining simply typed usage-aware syntaxes, and I give further non-$\name$
examples in \cref{sec:rec}.
Finally, I justify connections to linear logic and modal logic in
\cref{sec:translation}, where I translate $\name$ terms to and from
DILL~\citep{Barber1996} and the modal calculus of \citet{judgmental}.
%The application of the techniques of \cref{sec:simple} is left to
%\cref{sec:ren-sub-lr}.

\section{A usage-annotated calculus $\name$}\label{sec:lr}
\input{lr.tex}
\section{Bunched connectives}\label{sec:lnd}
\input{lnd.tex}
\section{Additions to and variations of \name{}}\label{sec:variant}
\input{rec.tex}
\section{Translation to and from existing systems}\label{sec:translation}
\input{trans.tex}

\section{Conclusion}\label{sec:semirings-conc}

In this chapter, I have presented the calculus $\name$, with particular focus on
the posemiring usage annotations.
Posemirings are sufficient for many use cases, as shown by the examples in this
chapter.
However, there are many more substructural disciplines in the literature which
cannot be expressed using posemirings.

An important point to note is that $\name$ cannot be instantiated to become a
bunched logic, in the sense of \citet{oHP99}.
This is despite the use of the bunched connectives in the \emph{definition} of
$\name$, and despite the formally tree-shaped contexts.
However, the variable rule of $\name$ essentially treats the variables in the
context independently, except for the individual checking of usage annotations
on each variable.
Therefore, we cannot talk about more interesting spatial relationships between
variables, as required by bunched logic.
This property also precludes us from capturing calculi without exchange
(non-commutative logics), such as Lambek calculus~\citep{Lambek58}.
Fitch-style systems\todo{cite} probably also are precluded similarly.

The work I present here has some methodological similarities with the earlier
work of \citet{LicataSR17}.
In that framework, one can provide a very precise \emph{mode theory}, expressing
which structural rules are available, and from it get a sequent calculus with
cut-elimination.
They encode many systems, including a non-associative logic (where context are
trees obeying no structural rules) and a bunched logic, and the framework is
probably expressive enough to encode all posemiring-based usage disciplines.
However, aside from the structural rules, the sequent calculus is fixed ---
there are two connectives: $\mathsf F$ (internalising the left of the sequent,
comparable to my
$\oc_{\gr{r_1}}\plr{-} \otimes \cdots \otimes \oc_{\gr{r_n}}\plr{-}$) and
$\mathsf U$ (internalising the whole sequent, comparable to my
$\oc_{\gr{r_1}}\plr{-} \multimap \cdots \multimap \oc_{\gr{r_n}}\plr{-}
\multimap \plr{-}$), which have left and right rules, and then the only other
rule is the variable rule.
This restriction is what allows the cut elimination theorem to be even
plausible, but means that other connectives, like additives and inductive types,
and more exotic syntaxes, like those I will present in
\cref{sec:other-syntaxes}, would have to be developed from scratch.

The distinction between sharing and separating conjunction of premises naturally
falls out of the posemiring approach to usage restrictions.
However, a similar distinction also appears in other approaches to linearity,
and perhaps in other substructural systems.
Indeed, in the paper that inspired the bunched connectives~\citep{RPKV20},
linearity is enforced in a similar style as in Yalla~\citep{laurent18}, using
lists which can be split.
It would be interesting to see future work on substructural systems work out the
appropriate premise connectives and define their systems from there, rather than
directly manipulating contexts.
It may also be possible to see future work profitably abstracting over the
posemiring annotations, requiring only that contexts form something like a
commutative $\mathrm{Rel}$-monoid supporting variable-binding and
variable access.

### 2023-02-01 tex/syntax.tex
[ writing ] more

Unlike the single judgement form of \name{} and standard simply typed
$\lambda$-calculi, the $\mu\tilde\mu$-calculus has three judgement forms:
terms, coterms, and commands.
Read logically, terms and coterms are seen to, respectively, prove and refute
propositions (types), while commands exhibit contradictions.
This means that the abstract \AgdaBound{Ty} in the generic framework is
instantiated to \AgdaDatatype{Conc} (for \emph{conclusion}) as below, with
\AgdaDatatype{Ty} not being exposed directly to the generic framework.
For now, I just consider multiplicative disjunction $\parr$ (\emph{par}) and
negation/duality, beside an uninterpreted base type.
These are enough to exhibit classical behaviour.

@@

The syntax comprises a \emph{cut} between a term and a coterm of the same type,
the eponymous $\mu$ and $\tilde\mu$ constructs for proof by contradiction, and
then term and coterm (introduction and elimination) forms for negation and
\emph{par}.

@@

@@

As in the other examples, I define a data type of rule labels
\AgdaDatatype{`LnL} before defining the type system \AgdaFunction{LnL}.
The system \AgdaFunction{LnL} has types \AgdaFunction{$\Sigma$Ty} and usage
annotations $\{\gr0, \gr1, \gr\omega\}$, as we have used before for linear
systems.

@@

\begin{figure}
  \begin{mathpar}
    \ebrule[comb]{%
      \hypo{I^{\sep}}
      \infer1[$I$i]{\vdashL I}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL I}
      \hypo{\sep}
      \hypo{\vdashL C}
      \infer3[$I$e]{\vdashL C}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL A}
      \hypo{\sep}
      \hypo{\vdashL B}
      \infer3[$\otimes$i]{\vdashL A \otimes B}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL A \otimes B}
      \hypo{\sep}
      \hypo{\gr1A, \gr1B \vdashL C}
      \infer3[$\otimes$e]{\vdashL C}
    }
    \and
    \ebrule[comb]{%
      \hypo{\gr1A \vdashL B}
      \infer1[$\multimap$i]{\vdashL A \multimap B}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL A \multimap B}
      \hypo{\sep}
      \hypo{\vdashL A}
      \infer3[$\multimap$e]{\vdashL B}
    }
    \and
    \ebrule[comb]{%
      \hypo{\Boxzpt\plr{\vdashC X}}
      \infer1[$F$i]{\vdashL FX}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL FX}
      \hypo{\sep}
      \hypo{\gr\omega X \vdashL C}
      \infer3[$F$e]{\vdashL C}
    }
    \and
    \ebrule[comb]{%
      \hypo{\Boxzpt\dot1}
      \infer1[$1$i]{\vdashC 1}
    }
    \and
    \text{(no \TirName{$1$e})}
    \and
    \ebrule[comb]{%
      \hypo{\Boxzpt(\vdashC X}
      \hypo{\dottimes}
      \hypo{\vdashC Y)}
      \infer3[$\times$i]{\vdashC X \times Y}
    }
  \end{mathpar}
  \caption{Linear/non-Linear Logic in bunched notation}\label{fig:LnL-bunched}
\end{figure}

@@

\section{A layer of syntax is functorial}\label{sec:functorial}

A basic property of the universe of syntaxes
is that every syntax supports a functorial action on subterms, realised by a
function \AgdaFunction{map-s}.
Its type says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{$\Psi$} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{$\Psi$}.
The notation
\AgdaBound{$\Psi$}\AgdaSpace{}\AgdaSymbol{.}\AgdaField{rel}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}\AgdaBound{Q}
means the same as $\grQ \leq \grP\gr\Psi$ from \cref{sec:ren-sub-lr}.
The new notation comes about because I state \AgdaBound{$\Psi$} as a total and
deterministic linear relation, rather than a linear map, for reasons related to
mechanisation beyond the scope of this thesis.

\ExecuteMetaData[\Maptex]{map-s-type}

This generality is needed because usage contexts change between
a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
variables newly bound in that subterm).
Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{$\Psi$} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
systems, do not involve any changes in the usage contexts.

The linearity of relation \AgdaBound{$\Psi$} is given by fields
\AgdaField{rel-0$_m$}, \AgdaField{rel-+$_m$}, and \AgdaField{rel-*$_m$} (with
the subscript-m being a mnemonic for \emph{module}, as opposed to scalar).
\todo{Refer to preliminaries}

\ExecuteMetaData[\Maptex]{map-p}

I have also extended \AgdaFunction{map-p} to handle the various
$\Box$-modalities described in \cref{sec:dup-lnl}.
The Agda code for this extension is not particularly readable, so I do not
include it in this document.
However, this extension is notable as the only part of the framework requiring
that the linear relation \AgdaBound{$\Psi$} be functional (i.e., total and
deterministic).

>>>>

Unlike the single judgement form of \name{} and standard simply typed
$\lambda$-calculi, the $\mu\tilde\mu$-calculus has three judgement forms:
terms (traditional notation: $\Gamma \vdash A \mid \Delta$), coterms
($\Gamma \mid A \vdash \Delta$), and commands ($\Gamma \vdash \Delta$).
Read logically, terms and coterms are seen to, respectively, prove and refute
propositions (types), while commands exhibit contradictions.
This means that the abstract \AgdaBound{Ty} in the generic framework is
instantiated to \AgdaDatatype{Conc} (for \emph{conclusion}) as below, with
\AgdaDatatype{Ty} not being exposed directly to the generic framework.
For now, I just consider multiplicative disjunction $\parr$ (\emph{par}) and
negation/duality, beside an uninterpreted base type.
These are enough to exhibit classical behaviour.

@@

The syntax comprises a \emph{cut} between a term and a coterm of the same type,
the eponymous $\mu$ and $\tilde\mu$ constructs for proof by contradiction, and
then term and coterm (introduction and elimination) forms for negation and
\emph{par}.
I give a traditional presentation of this fragment of the
$\mu\tilde\mu$-calculus in \cref{fig:mmt-orig}, with my encoding of the rules
below.

@@

\begin{figure}
  \begin{mathpar}
    \ebrule{%
      \hypo{\grP \leq \grPl + \grPr}
      \hypo{\grQ \leq \grQl + \grQr}
      \hypo{\grPl\gamma \vdash A \mid \grQl\delta}
      \hypo{\grPr\gamma \mid A \vdash \grQr\delta}
      \infer4[Cut$_A$]{\grP\gamma \vdash \grQ\delta}
    } \and
    \ebrule{%
      \hypo{\Gamma \vdash \gr1A, \Delta}
      \infer1[$\mu$]{\Gamma \vdash A \mid \Delta}
    } \and
    \ebrule{%
      \hypo{\Gamma, \gr1A \vdash \Delta}
      \infer1[$\tilde\mu$]{\Gamma \mid A \vdash \Delta}
    } \and
    \ebrule{%
      \hypo{\Gamma \mid A \vdash \Delta}
      \infer1[$\lambda$]{\Gamma \vdash A^\bot \mid \Delta}
    } \and
    \ebrule{%
      \hypo{\Gamma \vdash A \mid \Delta}
      \infer1[$\tilde\lambda$]{\Gamma \mid A^\bot \vdash \Delta}
    } \and
    \ebrule{%
      \hypo{\grP \leq \gr r\grPl + \gr s\grPr}
      \hypo{\grQ \leq \gr r\grQl + \gr s\grQr}
      \hypo{\grPl\gamma \vdash A \mid \grQl\delta}
      \hypo{\grPr\gamma \vdash B \mid \grQr\delta}
      \infer4[$\alr{-,-}$]%
      {\grP\gamma \vdash \gr rA \parr \gr sB \mid \grQ\delta}
    } \and
    \ebrule{%
      \hypo{\Gamma, \gr rA, \gr sB \vdash \Delta}
      \infer1[$\mu\alr{-,-}$]{\Gamma \mid \gr rA \parr \gr sB \vdash \Delta}
    }
  \end{mathpar}
  \caption{A fragment of a usage-annotated $\mu\tilde\mu$-calculus presented in
    traditional sequent notation}
  \label{fig:mmt-orig}
\end{figure}

@@

As in the other examples, I define a data type of rule labels
\AgdaDatatype{`LnL} before defining the type system \AgdaFunction{LnL}.
The system \AgdaFunction{LnL} has types \AgdaFunction{$\Sigma$Ty} and usage
annotations $\{\gr0, \gr1, \gr\omega\}$, as we have used before for linear
systems.
For reference, I give a standard presentation of the rules of L/nL in
\cref{fig:LnL-orig}

@@

\begin{figure}
  \begin{align*}
    \mathcal J \Coloneqq \; \Theta; \Gamma \vdashL A \; \mid \; \Theta \vdashC X
  \end{align*}
  \begin{mathpar}
    \ebrule{%
      \infer0[$\mathcal L$-var]{\Theta; A \vdashL A}
    } \and
    \ebrule{%
      \infer0[$\mathcal C$-var]{\Theta, X \vdashC X}
    } \and
    \ebrule{%
      \infer0[$I$i]{\Theta; {\cdot} \vdashL I}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL I}
      \hypo{\Theta; \Delta \vdashL C}
      \infer2[$I$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A}
      \hypo{\Theta; \Delta \vdashL B}
      \infer2[$\otimes$i]{\Theta; \Gamma, \Delta \vdashL A \otimes B}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A \otimes B}
      \hypo{\Theta; \Delta, A, B \vdashL C}
      \infer2[$\otimes$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma, A \vdashL B}
      \infer1[$\multimap$i]{\Theta; \Gamma \vdashL A \multimap B}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL A \multimap B}
      \hypo{\Theta; \Delta \vdashL A}
      \infer2[$\multimap$e]{\Theta; \Gamma, \Delta \vdashL B}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X}
      \infer1[$F$i]{\Theta; {\cdot} \vdashL FX}
    } \and
    \ebrule{%
      \hypo{\Theta; \Gamma \vdashL FX}
      \hypo{\Theta, X; \Delta \vdashL C}
      \infer2[$F$e]{\Theta; \Gamma, \Delta \vdashL C}
    } \and
    \ebrule{%
      \infer0[$1$i]{\Theta \vdashC 1}
    } \and
    \text{(no \TirName{$1$e})} \and
    \ebrule{%
      \hypo{\Theta \vdashC X}
      \hypo{\Theta \vdashC Y}
      \infer2[$\times$i]{\Theta \vdashC X \times Y}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X_0 \times X_1}
      \infer1[$\times$e$_i$]{\Theta \vdashC X_i}
    } \and
    \ebrule{%
      \hypo{\Theta, X \vdashC Y}
      \infer1[$\to$i]{\Theta \vdashC X \to Y}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC X \to Y}
      \hypo{\Theta \vdashC X}
      \infer2[$\to$e]{\Theta \vdashC Y}
    } \and
    \ebrule{%
      \hypo{\Theta; {\cdot} \vdashL A}
      \infer1[$G$i]{\Theta \vdashC GA}
    } \and
    \ebrule{%
      \hypo{\Theta \vdashC GA}
      \infer1[$G$e]{\Theta; {\cdot} \vdashL A}
    }
  \end{mathpar}
  \caption{Linear/non-Linear Logic in traditional sequent notation}
  \label{fig:LnL-orig}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \ebrule[comb]{%
      \hypo{I^{\sep}}
      \infer1[$I$i]{\vdashL I}
    } \and
    \ebrule[comb]{%
      \hypo{\vdashL I}
      \hypo{\sep}
      \hypo{\vdashL C}
      \infer3[$I$e]{\vdashL C}
    } \and
    \ebrule[comb]{%
      \hypo{\vdashL A}
      \hypo{\sep}
      \hypo{\vdashL B}
      \infer3[$\otimes$i]{\vdashL A \otimes B}
    } \and
    \ebrule[comb]{%
      \hypo{\vdashL A \otimes B}
      \hypo{\sep}
      \hypo{\gr1A, \gr1B \vdashL C}
      \infer3[$\otimes$e]{\vdashL C}
    } \and
    \ebrule[comb]{%
      \hypo{\gr1A \vdashL B}
      \infer1[$\multimap$i]{\vdashL A \multimap B}
    } \and
    \ebrule[comb]{%
      \hypo{\vdashL A \multimap B}
      \hypo{\sep}
      \hypo{\vdashL A}
      \infer3[$\multimap$e]{\vdashL B}
    } \and
    \ebrule[comb]{%
      \hypo{\Box\plr{\vdashC X}}
      \infer1[$F$i]{\vdashL FX}
    } \and
    \ebrule[comb]{%
      \hypo{\vdashL FX}
      \hypo{\sep}
      \hypo{\gr\omega X \vdashL C}
      \infer3[$F$e]{\vdashL C}
    } \and
    \ebrule[comb]{%
      \hypo{\Box\dot1}
      \infer1[$1$i]{\vdashC 1}
    } \and
    \text{(no \TirName{$1$e})} \and
    \ebrule[comb]{%
      \hypo{\Box(\vdashC X}
      \hypo{\dottimes}
      \hypo{\vdashC Y)}
      \infer3[$\times$i]{\vdashC X \times Y}
    } \and
    \ebrule[comb]{%
      \hypo{\Box\plr{\vdashC X_0 \times X_1}}
      \infer1[$\times$e$_i$]{\vdashC X_i}
    } \and
    \ebrule[comb]{%
      \hypo{\Box\plr{\gr\omega X \vdashC Y}}
      \infer1[$\to$i]{\vdashC X \to Y}
    } \and
    \ebrule[comb]{%
      \hypo{\Box(\vdashC X \to Y}
      \hypo{\dottimes}
      \hypo{\vdashC X)}
      \infer3[$\to$e]{\vdashC Y}
    } \and
    \ebrule[comb]{%
      \hypo{\Box\plr{\vdashL A}}
      \infer1[$G$i]{\vdashC GA}
    } \and
    \ebrule[comb]{%
      \hypo{\Box\plr{\vdashC GA}}
      \infer1[$G$e]{\vdashL A}
    }
  \end{mathpar}
  \caption{Linear/non-Linear Logic in bunched notation using
    $\{\gr0, \gr1, \gr\omega\}$ usage annotations and with $\Boxzpt$ abbreviated
    to $\Box$}
  \label{fig:LnL-bunched}
\end{figure}

@@

\section{Conclusion}

I have shown via examples that the syntax descriptions defined in this chapter
can capture a significant range of simply typed usage-aware calculi.
Together with the examples of posemirings given in \cref{sec:semirings}, the
whole framework is capable of encoding many specific substructural calculi.
However, the language of descriptions has several restrictions, with more and
less clear practical consequences.
Many of these restrictions are shared with the syntax descriptions of
\citet{AACMM21}, but some pertain to usage annotations.

One clear limitation is that the forms of judgements do not allow enough
interdependency to capture dependently typed calculi.
In particular, for any given calculus, there is one fixed set of types, with no
indexing over contexts.
This allows contexts to have a flat list-like structure rather than a telescope,
and means that the type of a term does not depend on that context, but clearly
restricts the expressiveness of the calculi which can be encoded.
As well as dependently typed calculi, this also prevents the encoding of the
well formed types of System $F_\omega$.

Less clear than describability of System $F_\omega$ is describability of System
$F$.
Though I have not presented the details, I believe it should be possible to
encode calculi with polymorphism, such as System $F$ and the usage-annotated
system found in \citet{AbelBernardy2020}, via a two-stage process.
First, we make a syntax description for the language of types, and then use
the well typed terms arising from that description as the set of types when
making the description for the System $F$ terms.
Such an encoding would yield all of the benefits of the framework presented in
this thesis separately to the language of types and the language of terms, but
how these two languages interact is unclear.
For example, in \cref{sec:semantics}, I provide a syntax-generic substitution
operation.
This operation would yield substitution for type variables in types and
substitution for term variables in terms, but substitution for type variables in
terms would have to be provided separately, and would likely be more
complicated.
It would be interesting to see future work in this direction, possibly inspired
by how Autosubst~\citep{Autosubst15} handles the same issue.

As for restrictions peculiar to the usage-aware setting, we of course have the
general limitations of posemiring annotations listed in
\cref{sec:semirings-conc}, which I will not repeat here.
Additionally, there are things we could wish to do with posemiring-based usage
annotations which we do not have access to with the descriptions given in this
chapter.
For example, an issue one may have with the encoding of Linear/non-Linear Logic
given in \cref{sec:dup-lnl} is that linear variables and intuitionistic
variables, despite having morally distinct usage disciplines, necessarily share
the $\{\gr0, \gr1, \gr\omega\}$ posemiring of usage annotations.
If one wanted to generalise this construction to different pairs of usage
disciplines, one may well find that they do not combine adequately into a single
posemiring.
A possible solution would be to have a range of kinds, and for each kind to have
its own set of types and its own posemiring --- for example, a \emph{linear}
kind and an \emph{intuitionistic} kind, using $\{\gr0, \gr1, \gr\omega\}$ and
the 1-element posemiring, respectively.
However, the descriptions given in this thesis only allow for one posemiring.

Finally, while there is a simple and well motivated core of premise combinators
--- $\dot1$, $\dottimes$, $I^{\sep}$, $\sep$, and $\gr r \cdot \plr{-}$ --- the
various $\Box$-modalities feel like ad hoc additions.
While I have shown that $\Boxzp$ and $\Boxzpt$ have several practical uses, it
is unclear whether there should be more modalities.
Furthermore, I have not made an effort to isolate the properties that make these
modalities well behaved.
We will see in \cref{sec:functorial} that any such modality will need to be
functorial in the appropriate sense, but other examples of generic programming,
such as the usage elaborator which will be detailed in
\cref{sec:usage-elaborator}, may need further properties.
